Supplemental Information: Gas exchange in large rivers controlled by largest turbulent eddies: implications for remotely sensing gas exchange via SWOT

### 2021-07-26 supp_v4.Rmd
rewriting

---
title: "Gas exchange in big rivers controlled by stream power and turbulence scales: implications for remotely sensing gas exchange"
author: "CB Brinkerhoff, CJ Gleason, PA Raymond, MH Harlan"
date: "`r Sys.Date()`"
csl: agu_citation_style.csl
output:
  word_document: 
      reference_docx: word-styles-reference.docx
bibliography: References.bib

@@

## Text S1: Deriving two expressions for turbulent kinetic energy production
Below we detail how we arrived at our two expressions for a cross-section-averaged turbulent kinetic energy production (*G*), or equation 2 in the main text.

@@

$A_0$ prior hyperparameters were assigned following an updated version of the method developed by @brinkerhoffConstrainingRemoteRiver2020a. They developed a set of river channel prior hyperparameters for McFLI algorithms that are entirely RS-able and reflect differential channel hydraulics as a function of river geomorphology. They used an extensive database of field measurements and machine learning to identify patterns that associate river width with the hydraulic priors needed to run McFLIs so that prior hyperparameters may be assigned to rivers using only the existing remotely sensed data. For this study, we extracted $\lambda$ and $\gamma$ as the Xth percentile values rather than the absoulte maximum and minimum values to avoid physically impossible bounds on $A_0$. This leaves the $k_{600}$ hyperparameters to be defined. We assigned those by using a simple slope regression model to predict $U_*$, trained on the @ulsethDistinctAirWater2019a: equation S9 assigns $\mu$ while $\sigma^2$ was set to equation S9's log-transformed standard error after propogation of errors for both equation S9 and equation 4 in the main text: 1.12. $\lambda$ and $\gamma$ were set to log(0.001) m/day and log(500) m/day, respectively. Finally, we estimate $\sigma_{k_{600}}$ using the standard model error from equation 4 in the main text and as fit on the data shown in Figure 3a.

@@

## Text S3: BIKER validation setup
Regardless of the validation setup or SWOT error budget used, we do not have observed $k_{600}$ data for these rivers, and to our knowledge no field dataset of $k_{600}$ exists in the type of temporal and spatial frequency that SWOT (and therefore the BIKER algorithm) provides. Therefore, we take the model outlined in row 1 of Table 1 and use that to calculate the observed $k_{600}$, that BIKER is validated against (equation S10).

@@

It also means that, for a fair validation scheme, $\sigma_{k_{600}}^2$ in equation 5 in the main text must be set to reflect only error from our assumptions about calculating *dA* and not the parameter uncertainity inherent in the model coefficient (i.e. the 56.0294 in equation 4 in the main text). Because this model is only theoretically valid when $R_h=H$, we convientantly do not need to account for uncertanties associated with assuming that $R_h = H$. So for this validation, $\sigma_{k_{600}}^2$ is set to 0.20 (natural-log space) to reflect a small degree of uncertainity that could arise from assuming a rectangular river channel. However once SWOT launches, it should reflect the total uncertainty described in Text S3 (1.12).

Validation is performed using the BIKER posterior means. Validation metrics take two forms (and are detailed in Table S1). To validate across all rivers and timesteps, we used the coefficient of determination $r^2$ and the root mean square error RMSE. Three normalized metrics were used for by-river validation: RRMSE and NRMSE are normalized root mean square errors that have been normalized by the observed value and the mean observed value (respectively). rBIAS is a measure of prediction bias that is normalized by the mean observed value. r2 is again used on the by-river case.

## Text S4: BIKER FCO2 calculations
To calculate $CO_2$ fluxes and carbon efflux, we pair @beaulieuControlsGasTransfer2012a's biweekly 26 $CO_2$ and water temperature samples with every 11th SWOT observation by date, ignoring the timesteps beyond 26. This amounts to only ~15% of the SWOT observations and we deem this acceptable. We sample every 11 days as this is the average sampling resolution for SWOT and the $CO_2$ and water temperature data are approximately the same at a 14 day resolution. Not all of the SWOT rivers have observations for a full year, and when simulation dates are not available they are assumed to start on January 1st. We then pair the modeled $k_{CO_2}$ values (obtained from $k_{600}$ and equation S11) with these water-side $CO_2$ concentrations and water temperatures. In equation S11, *Sc* is the Schmidt number at the observed water temperature. This was calculated following @raymondScalingGasTransfer2012a and @wanninkhofRelationshipWindSpeed1992a. Atmospheric $CO_2$ was assumed 390 uatm. $FCO_2$ validation was performed using the same metrics as $k_{600}$ validation (Table 1).

@@

![Figure S3 Stuff](cache/k600_theory/swotPlot.jpg)

>>>>

---
title: "Supplemental Information: Gas exchange in large rivers controlled by largest turbulent eddies: implications for remotely sensing gas exchange via SWOT"
author: "CB Brinkerhoff, CJ Gleason, PA Raymond, MH Harlan"
date: "`r Sys.Date()`"
csl: agu_citation_style.csl
output:
  word_document: 
      reference_docx: word-styles-reference.docx
bibliography: References.bib

@@

## Text S1: Deriving two expressions for turbulent kinetic energy production
Below we detail how we arrived at our two expressions for a cross-section-averaged turbulent kinetic energy production (*G*), or equation 3 in the main text. Please see Appendix A for variable descriptions and notation.

@@

$A_0$ prior hyperparameters were assigned following an updated version of the method developed by @brinkerhoffConstrainingRemoteRiver2020a. They developed a set of river channel prior hyperparameters for McFLI algorithms that are entirely RS-able and reflect differential channel hydraulics as a function of river geomorphology. They used an extensive database of field measurements and machine learning to identify patterns that associate river width with the hydraulic priors needed to run McFLIs so that prior hyperparameters may be assigned to rivers using only the existing remotely sensed data. For this study, we extracted $\lambda_{A_0}$ and $\gamma_{A_0}$ as the **Xth** percentile values rather than the absolute maximum and minimum values to avoid physically impossible bounds on $A_0$.

This leaves the $k_{600}$ hyperparameters to be defined. We assigned those by using a simple slope regression model to predict $U_*$ solely from a SWOT-observable value. It was trained on the @ulsethDistinctAirWater2019a dataset. Equation S9 assigns $\mu_{k_{600}}$ while $\sigma^2_{k_{600}}$ was set to equation S9's log-transformed standard error after propogation of errors for both equation S9 and equation 6 in the main text. This amounted to a value of 1.12. $\lambda_{k_{600}}$ and $\gamma_{k_600}}$ were set to log(0.001) m/day and log(500) m/day, respectively. Finally, we estimate $\sigma_{k_{600}}$ using the standard model error from equation 6 in the main text and as fit on the data shown in Figure 3a.

@@

## Text S3: BIKER validation setup
Regardless of the validation setup or SWOT error budget used, we do not have observed $k_{600}$ data for these rivers, and to our knowledge no field dataset of $k_{600}$ exists in the type of temporal and spatial frequency that SWOT (and therefore the BIKER algorithm) provides. Therefore, we take the model outlined in row 1 of Table 1 in the main text and use that to calculate the observed $k_{600}$, that BIKER is validated against (equation S10).

@@

It also means that, for a fair validation scheme, $\sigma_{k_{600}}^2$ in equation 7 in the main text must be set to reflect only error from our assumptions about calculating *dA* and not the parameter uncertainity inherent in the model coefficient (i.e. the 56.0294 in equation 6 in the main text). Because this model is only theoretically valid when $R_h=H$, we convientantly do not need to account for uncertanties associated with assuming that $R_h = H$. So for this validation, $\sigma_{k_{600}}^2$ is set to 0.20 (natural-log space) to reflect a small degree of uncertainity that could arise from assuming a rectangular river channel. However once SWOT launches, it should reflect the total uncertainty described in Text S3 (1.12).

Validation is performed using the BIKER posterior means. Validation metrics take two forms (and are detailed in Table S1). To validate across all rivers and timesteps, we used the coefficient of determination $r^2$ and the root mean square error RMSE. Four normalized metrics were used for by-river validation: RRMSE and NRMSE are normalized root mean square errors that have been normalized by the observed value and the mean observed value (respectively). rBIAS is a measure of prediction bias that is normalized by the mean observed value. r2 is again used on the by-river case too.

## Text S4: BIKER FCO2 workflow
To calculate $CO_2$ fluxes and carbon efflux, we pair @beaulieuControlsGasTransfer2012a's biweekly 26 $CO_2$ and water temperature samples with every 11th SWOT observation by date, ignoring the timesteps beyond 26. This amounts to only ~15% of the SWOT observations and we deem this acceptable. We sample every 11 days as this is the average sampling resolution for SWOT and the $CO_2$ and water temperature data are approximately the same at a 14 day resolution. Not all of the SWOT rivers have observations for a full year, and when simulation dates are not available they are assumed to start on January 1st. We then pair the modeled $k_{CO_2}$ values (obtained from $k_{600}$ and equation S11) with these water-side $CO_2$ concentrations and water temperatures. In equation S11, *Sc* is the Schmidt number at the observed water temperature. This was calculated following @raymondScalingGasTransfer2012a and @wanninkhofRelationshipWindSpeed1992a. Atmospheric $CO_2$ was assumed 390 uatm. $FCO_2$ validation was performed using the same metrics as $k_{600}$ validation (Table 1).

@@

![Figure S3 Cummulative density functions (CDFs) of river width measurements, as binned by their dataset and and whether or not they meet the 'large river' condition (Rh=H). The dashed black line indicates the minimum river width that the SWOT satellite will observe (100m). This brief analysis indicates that the vast majority of SWOT-observable rivers should meet the 'large river' condition.](cache/k600_theory/swotPlot.jpg)
